---
title: "Week 05 Questions"
author: "Scott Graham 10131323"
date: "October 13, 2017"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse, warn.conflicts = FALSE, quietly = TRUE)
library(magrittr, warn.conflicts = FALSE, quietly = TRUE)
theme_minimal2 <- theme_minimal() %>%  theme_set()
theme_minimal2 <-
  theme_update(
    panel.border = element_rect(
      linetype = "solid"
      ,colour = "grey92"
      ,fill = NA
    )
    ,strip.background = element_rect(
      linetype = "solid"
      ,colour = "grey92"
      ,fill = NA
    )
  )
```
\[
  \newcommand{\Prob}{\operatorname{P}}
  \newcommand{\E}{\operatorname{E}}
  \newcommand{\Var}{\operatorname{Var}}
  \newcommand{\Cov}{\operatorname{Cov}}
  \newcommand{\se}{\operatorname{se}}
  \newcommand{\re}{\operatorname{re}}
  \newcommand{\ybar}{{\overline{Y}}}
  \newcommand{\phat}{{\hat{p}}}
  \newcommand{\that}{{\hat{T}}}
  \newcommand{\med}{{\tilde{Y}}}
\]

## 4.9
```{r 4.09}
crabs_data <- 
  read_delim(
    file = "http://www.math.montana.edu/shancock/courses/stat539/data/horseshoe.txt"
    ,delim = " "
    ,col_names = TRUE
  ) %>% 
  mutate(
    satell_bool = if_else(satell > 0, 1, 0)
    ,c1 = if_else(color == 1, 1, 0 )
    ,c2 = if_else(color == 2, 1, 0 )
    ,c3 = if_else(color == 3, 1, 0 )
    ,color = color + 1
  )
```
### a.
```{r 4.09a}
logit_reg_4.09a <- glm(satell_bool ~ c1 + c2 + c3, data = crabs_data, family = "binomial")
summary(logit_reg_4.09a)
```
$$
  \ln\left( \theta \right) =
  -0.7621 + 1.8608c_{1} + 1.7382c_{2} + 1.1299c_{3}
$$
That is, on average for a medium light crab, the log odds in favour increase by 1.8608 compared to a dark crab, based on the sample. Or the average odds in favour increase by a multiplicative factor of $e^{1.8608} = 6.4286$, based on the sample.

### b.
```{r 4.09b}
pchisq(
  q = logit_reg_4.09a$null.deviance - logit_reg_4.09a$deviance
  ,df = logit_reg_4.09a$df.null - logit_reg_4.09a$df.residual
  ,lower.tail = FALSE
)
```
We therefore reject our null hypothesis that the models have similar levels of deviance, and accept the alternative that they differ, based on the sample. That is, colour does play a role in predicting if a crab has satellites.

### c.
```{r 4.09c}
logit_reg_4.09c <- glm(satell_bool ~ color, data = crabs_data, family = "binomial")
summary(logit_reg_4.09c)
```
$$
  \ln\left( \theta \right) =
  3.0781 - 0.7147Colour
$$
As you move each step up the colour scale, with 2 being medium light, and 5 being dark, on average the log odds in favour decrease by 0.7147, or the odds decrease by a multiplicative factor of $e^{-0.7147} = 0.4893$, based on the sample.

### d.
```{r 4.09d}
pchisq(
  q = logit_reg_4.09c$null.deviance - logit_reg_4.09c$deviance
  ,df = logit_reg_4.09c$df.null - logit_reg_4.09c$df.residual
  ,lower.tail = FALSE
)
```
Therefore we reject the null hypothesis that the models explain the same amount of variance, and accept that the more complicated model is better based on the sample.

### e.
The advantage of quantitative colour, is we set the $df$ in the likelihood ratio test to 1, giving a more powerful test, but if there is a non-linear relationship between colour and the log odds in favour (polynomial, exponential, etc), this model may not capture it.


## 4.11
$$
  \frac{\theta_{\text{Spouse}}}{\theta_{\text{Alone}}} =
  2.02,
  \frac{\theta_{\text{Others}}}{\theta_{\text{Alone}}} =
  1.71 \implies
  \frac{\theta_{\text{Spouse}}}{\theta_{\text{Others}}} =
  \frac{2.02}{1.71} =
  1.18
$$
$$
  \frac{\theta_{\text{Medium Income}}}{\theta_{\text{Low Income}}} =
  0.72,
  \frac{\theta_{\text{High Income}}}{\theta_{\text{Low Income}}} =
  0.41 \implies
  \frac{\theta_{\text{High Income}}}{\theta_{\text{Medium Income}}} =
  \frac{0.41}{0.72} =
  0.57
$$


## 4.16
```{r 4.16}
myers_data <- 
  tibble(
    Personality = c(
      rep("ESTJ", times = 10 + 67)
      ,rep("ESTP", times = 8 + 34)
      ,rep("ESFJ", times = 5 + 101)
      ,rep("ESFP", times = 7 + 72)
      ,rep("ENTJ", times = 3 + 20)
      ,rep("ENTP", times = 2 + 16)
      ,rep("ENFJ", times = 4 + 27)
      ,rep("ENFP", times = 15 + 65)
      ,rep("ISTJ", times = 17 + 123)
      ,rep("ISTP", times = 3 + 49)
      ,rep("ISFJ", times = 6 + 132)
      ,rep("ISFP", times = 4 + 102)
      ,rep("INTJ", times = 1 + 12)
      ,rep("INTP", times = 5 + 30)
      ,rep("INFJ", times = 1 + 30)
      ,rep("INFP", times = 6 + 73)
    )
    ,Alcohol = c(
      rep(c(1, 0), times = c(10, 67))
      ,rep(c(1, 0), times = c(8, 34))
      ,rep(c(1, 0), times = c(5, 101))
      ,rep(c(1, 0), times = c(7, 72))
      ,rep(c(1, 0), times = c(3, 20))
      ,rep(c(1, 0), times = c(2, 16))
      ,rep(c(1, 0), times = c(4, 27))
      ,rep(c(1, 0), times = c(15, 65))
      ,rep(c(1, 0), times = c(17, 123))
      ,rep(c(1, 0), times = c(3, 49))
      ,rep(c(1, 0), times = c(6, 132))
      ,rep(c(1, 0), times = c(4, 102))
      ,rep(c(1, 0), times = c(1, 12))
      ,rep(c(1, 0), times = c(5, 30))
      ,rep(c(1, 0), times = c(1, 30))
      ,rep(c(1, 0), times = c(6, 73))
    )
  ) %>% 
  # If the stated letter appears in the personality column, place a 1, 
  # otherwise place a 0.
  mutate(
    `E/I` = if_else(grepl(pattern = "E", x = Personality), 1, 0)
    ,`S/N` = if_else(grepl(pattern = "S", x = Personality), 1, 0)
    ,`T/F` = if_else(grepl(pattern = "T", x = Personality), 1, 0)
    ,`J/P` = if_else(grepl(pattern = "J", x = Personality), 1, 0)
  )
myers_data
```
### a.
```{r 4.16a}
logit_reg_4.16a <- glm(
  Alcohol ~ `E/I` + `S/N` + `T/F` + `J/P`
  ,data = myers_data
  ,family = "binomial"
)
summary(logit_reg_4.16a)
```
$$
  \ln\left( \theta \right) =
  -2.4668 + 0.5550\text{E/I} - 0.4292\text{S/N} + 0.6873\text{T/F} - 0.2022\text{J/P}
$$
Where the first letter for each coefficient is coded as a 1, and the second letter as a 0.

### b.
```{r 4.16b}
predict(
  object = logit_reg_4.16a
  ,newdata = 
    tibble(
    `E/I` = 1
    ,`S/N` = 1
    ,`T/F` = 1
    ,`J/P` = 1
  )
  ,type = "response")
```

### c.
```{r 4.16c}
predict(
  object = logit_reg_4.16a
  ,newdata = 
    tibble(
    `E/I` = 1
    ,`S/N` = 0
    ,`T/F` = 1
    ,`J/P` = 0
  )
  ,type = "response")
```
It is the response that has all the positive coefficients set to 1, and the negative coefficients set to 0.


## 4.17
### a.
```{r 4.17a}
logit_reg_4.17a <- glm(
  Alcohol ~ `E/I` + `T/F`
  ,data = myers_data
  ,family = "binomial"
)
summary(logit_reg_4.17a)
predict(
  object = logit_reg_4.17a
  ,newdata = 
    tibble(
    `E/I` = 0
    ,`T/F` = 0
  )
  ,type = "response")
```

### b.
$$
  e^{\beta_{\text{E/I}}} =
  e^{0.5805} =
  1.7869
$$
The odds in favour on average increases by a mulitplicative factor 1.7869 for extroverted personalities vs. introverted personalities, based on the sample.

### c.
```{r 4.17c}
confint(object = logit_reg_4.17a, level = 0.95) %>% 
  exp %>% 
  round(4)
```
With 95\% confidence, the true value of the conditional odds ratio between E/I and the response lies between 1.1722 and 2.7400 based on the sample.

### d.
$$
  \beta_{\text{I/E}} =
  -\beta_{\text{E/I}} =
  -0.5805 \implies
  e^{\beta_{\text{I/E}}} =
  \frac{1}{e^{\beta_{\text{E/I}}}} =
  e^{-0.5805} =
  0.5596
$$
95\% CI for the conditional odds ratio between I/E and the response:
$$
  \left[\frac{1}{2.7400}, \frac{1}{1.1722} \right] =
  [0.3650, 0.8531]
$$
With 95\% confidence, the true value of the conditional odds ratio between I/E and the response lies between 0.3650 and 0.8531 based on the sample.

### e.
```{r 4.17e}
logit_reg_4.17e <- glm(
  Alcohol ~ `T/F`
  ,data = myers_data
  ,family = "binomial"
)
summary(logit_reg_4.17e)
anova(logit_reg_4.17e, logit_reg_4.17a, test = "Chisq")
```
Using the likelihood ratio test, comparing the two models, a p-value of 0.006955 is observed. So at $\alpha=0.05$ the null hypothesis of equivalent explained variance is rejected, and it can be assumed that E/I plays a statistically significant effect based on the sample.


## 4.19
```{r 4.19}
logit_reg_4.19 <- function(Gender, Religion, Political){
  log_odds <-
    -0.11 + 
    if_else(Gender == "M", 0.0, 0.16) +
    if_else(Religion == "J", 0.0, if_else(Religion == "C", -0.66, -0.57)) +
    if_else(Political == "I", 0.0, if_else(Political == "R", -1.67, 0.84))
  log_odds
}
```
### a.
$$
  OR_{\text{Gender}} =
  e^{\hat{\beta}_{1}^{G} - \hat{\beta}_{2}^{G}} =
  e^{0.16 - 0} = 
  1.1735
$$
Holding all else equal, the odds of supporting legalized abortion increases on average by a multiplicative factor of 1.1735 for females vs. males.

### b.
```{r 4.19b}
log_odds_MCR <- logit_reg_4.19(Gender = "M", Religion = "C", Political = "R")
prob_MCR <- exp(log_odds_MCR)/(1 + exp(log_odds_MCR))
prob_MCR
log_odds_FJD <- logit_reg_4.19(Gender = "F", Religion = "J", Political = "D")
prob_FJD <- exp(log_odds_FJD)/(1 + exp(log_odds_FJD))
prob_FJD
```

### c.
$$
  \hat{\beta}_{2}^{G} = 
  0.16 \implies
  OR_{\text{Gender}} =
  e^{\hat{\beta}_{1}^{G} - \hat{\beta}_{2}^{G}} =
  e^{0 - 0.16} =
  0.8521
$$
Holding all else equal, the odds of supporting legalized abortion decreases on average by a multiplicative factor of 0.8521 for males vs. females.

### d.
$$
  \hat{\beta}_{1}^{G} + \hat{\beta}_{2}^{G} =
  0,
  \hat{\beta}_{1}^{G} - \hat{\beta}_{2}^{G} =
  0.16 \implies
  \hat{\beta}_{1}^{G} =
  0.08, 
  \hat{\beta}_{2}^{G} =
  -0.08 \implies
$$
$$
  OR_{\text{Gender}} =
  e^{0.08-(-0.08)} =
  e^{0.16} =
  1.1735
$$


## 4.21
### a.
The odds in favour of getting condoms increases on average by a multiplicative factor of 4.04 for those who wached the video, vs. those who didn't based on the sample. With 95\% confidence the true value of said odds ratio is between 1.17 and 13.9.

### b.
$$
  \ln\left( \theta_{\text{Condoms}} \right) =
  \hat{\beta_{0}} + \ln 4.04x_{\text{Group}} + \ln 1.38x_{\text{Gender}} + \ln 5.82x_{\text{SES}} + \ln 3.22x_{\text{Num. of Partners}} \implies
$$
$$
  \ln\left( \theta_{\text{Condoms}} \right) =
  \hat{\beta_{0}} + 1.40x_{\text{Group}} + 0.32x_{\text{Gender}} + 1.76x_{\text{SES}} + 1.17x_{\text{Num. of Partners}}
$$

### c.
$$
  \hat{\beta}_{\text{Gender}} =
  \frac{\ln 1.23 + \ln 12.88}{2} =
  1.3813 \approx
  1.38 \neq
  \ln 1.38 =
  0.3221 \implies 
$$
$$
  OR_{\text{Gender}} =
  e^{1.38} =
  3.9803 \approx
  3.98
$$
Therefore, if the confidence interval is correct the log odds ratio for gender must be 1.38, giving an estimated odds ratio of 3.98.