---
title: "Week 08 Questions"
author: "Scott Graham"
date: "November 3, 2017"
output:
  html_document:
    theme: lumen
    toc: true
    toc_depth: 2
    toc_float: true
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse, warn.conflicts = FALSE, quietly = TRUE)
library(magrittr, warn.conflicts = FALSE, quietly = TRUE)
library(flatr, warn.conflicts = FALSE, quietly = TRUE)
theme_minimal2 <- theme_minimal() %>%  theme_set()
theme_minimal2 <-
  theme_update(
    panel.border = element_rect(
      linetype = "solid"
      ,colour = "grey92"
      ,fill = NA
    )
    ,strip.background = element_rect(
      linetype = "solid"
      ,colour = "grey92"
      ,fill = NA
    )
  )
```
\[
  \newcommand{\Prob}{\operatorname{P}}
  \newcommand{\E}{\operatorname{E}}
  \newcommand{\Var}{\operatorname{Var}}
  \newcommand{\Cov}{\operatorname{Cov}}
  \newcommand{\se}{\operatorname{se}}
  \newcommand{\re}{\operatorname{re}}
  \newcommand{\ybar}{{\overline{Y}}}
  \newcommand{\phat}{{\hat{p}}}
  \newcommand{\that}{{\hat{T}}}
  \newcommand{\med}{{\tilde{Y}}}
\]

## 4.29
### Data
```{r 4.29 Data}
sex_ct <-
  array(
    data =
      c(
         43, 26, 134, 149
        ,29, 22, 23, 36
      )
    ,dim = c(2, 2, 2)
    ,dimnames = list(
      Gender = c("M", "F")
      ,Intercourse = c("Y", "N")
      ,Race = c("W", "B")
    )
  )
sex_ct

sex_ct_flip <-
  array(
    data =
      c(
         43, 29, 134, 23
        ,26, 22, 149, 36
      )
    ,dim = c(2, 2, 2)
    ,dimnames = list(
      Race = c("W", "B")
      ,Intercourse = c("Y", "N")
      ,Gender = c("M", "F")
    )
  )
sex_ct_flip

sex_data <- 
  tibble(
    Race = 
      as.factor(
        rep(
          c("W", "B")
        ,times = 
          c(
            sum(sex_ct[,,"W"])
            ,sum(sex_ct[,,"B"])
          )
        )
      )
    ,Gender =
      as.factor(
        rep(
          rep(c("M", "F"), times = dim(sex_ct)[3])
          ,times =
            c(
              sum(sex_ct["M",,"W"])
              ,sum(sex_ct["F",,"W"])
              ,sum(sex_ct["M",,"B"])
              ,sum(sex_ct["F",,"B"])
            )
        )
      )
    ,Intercourse =
      as.numeric(
        rep(
          rep(c(1, 0), times = dim(sex_ct)[3]*dim(sex_ct)[1])
          ,times =
            c(
              sum(sex_ct["M","Y","W"])
              ,sum(sex_ct["M","N","W"])
              ,sum(sex_ct["F","Y","W"])
              ,sum(sex_ct["F","N","W"])
              ,sum(sex_ct["M","Y","B"])
              ,sum(sex_ct["M","N","B"])
              ,sum(sex_ct["F","Y","B"])
              ,sum(sex_ct["F","N","B"])
            )
        )
      )
  )
sex_summary <- 
  sex_data %>% 
  group_by(Race, Gender) %>% 
  summarize(
    Yes = sum(Intercourse)
    ,No = length(Intercourse) - Yes
    ,Total = Yes + No
  )
sex_summary

sex_plot_data <- 
  sex_summary %>%
  ungroup %>% 
  mutate(
    Race = if_else(Race == "W", "White", "Black")
    ,Gender = if_else(Gender == "M", "Male", "Female")
    ,Yes = Yes/Total
    ,No = 1 - Yes
  ) %>% 
 dplyr::select(
    -Total
  ) %>%
  gather(
    key = Intercourse
    ,value = `Sample Proportion`
    ,-Race
    ,-Gender
  )

sex_plot_data %>% 
  ggplot(
    aes(
      x = Gender
      ,y = `Sample Proportion`
      ,fill = Intercourse
    )
  ) +
  geom_col(alpha = 0.75) +
  facet_wrap(
    ~ Race
    ,nrow = 1
  ) +
  scale_fill_brewer(
    type = "qual"
    ,name = "Intercourse:"
    ,palette = "Set2"
  ) +
  labs(
    title = "Proportion of Persons who have had Intercourse by Gender"
    ,subtitle = "Conditional on Race"
  )

sex_plot_data %>% 
  ggplot(
    aes(
      x = Race
      ,y = `Sample Proportion`
      ,fill = Intercourse
    )
  ) +
  geom_col(alpha = 0.75) +
  facet_wrap(
    ~ Gender
    ,nrow = 1
  ) +
  scale_fill_brewer(
    type = "qual"
    ,name = "Intercourse:"
    ,palette = "Set2"
  ) +
  labs(
    title = "Proportion of Persons who have had Intercourse by Race"
    ,subtitle = "Conditional on Gender"
  )
```

### Cochran-Mantel-Haenszel Test
$$
  H_{0}:\text{ Intercourse is independent of Gender, conditional on Race}
$$
$$
  H_{1}:\text{ Intercourse is dependent on Gender, conditional on Race}
$$
```{r 4.29 CMH}
sex_cmh <- mantelhaen.test(sex_ct, correct = FALSE)
sex_cmh
```
With a p-value of `r round(sex_cmh$p.value, 4)`, we reject the null hypothesis, and accept the alternative based on the sample.

$$
  H_{0}:\text{ Intercourse is independent of Race, conditional on Gender}
$$
$$
  H_{1}:\text{ Intercourse is dependent on Race, conditional on Gender}
$$
```{r 4.29 CMH Flip}
sex_cmh_flip <- mantelhaen.test(sex_ct_flip, correct = FALSE)
sex_cmh_flip
```
With a p-value of `r round(sex_cmh_flip$p.value, 4)`, we reject the null hypothesis, and accept the alternative based on the sample.

### Logistic Regression
```{r 4.29 Logit}
logit_sex_01 <- 
  sex_data %>% 
  glm(
    Intercourse ~ Race + Gender + Race:Gender
    ,family = binomial
    ,data = .
  )
summary(logit_sex_01)

logit_sex_02 <-
  sex_data %>% 
  glm(
    Intercourse ~ Race + Gender
    ,family = binomial
    ,data = .
  )
summary(logit_sex_02)

anova(logit_sex_02, logit_sex_01, test = "Chisq")
```
By doing a Likelihood Ratio Test between the model with an interaction term, and the model without, a p-value of `r round(anova(logit_sex_02, logit_sex_01, test = "Chisq")$"Pr(>Chi)"[2], 4)` is found. This fails to reject the null hypothesis of explaining equal amounts of variance, based on the sample. As such we can use the more parsimonious model.

Both the slope parameters are significant at $\alpha=0.05$, based on the sample. A White person on average has an increased log likelihood in favour of having intercourse of `r round(coefficients(logit_sex_02)[2], 4)` vs. a Black person, based on the sample. A Male on average has an increased log likelihood in favour of having intercourse of `r round(coefficients(logit_sex_02)[3], 4)` vs. a Female, based on the sample.


## 4.31
### Data
```{r 4.31 Data}
marijuana_ct <-
  array(
    data =
      c(
         911, 44, 538, 456
        ,3, 2, 43, 279
      )
    ,dim = c(2, 2, 2)
    ,dimnames = list(
      Cigarette = c("Y", "N")
      ,Marijuana = c("Y", "N")
      ,Alcohol = c("Y", "N")
    )
  )
marijuana_ct

marijuana_ct_flip <-
  array(
    data =
      c(
         911, 3, 538, 43
        ,44, 2, 456, 279
      )
    ,dim = c(2, 2, 2)
    ,dimnames = list(
      Alcohol = c("Y", "N")
      ,Marijuana = c("Y", "N")
      ,Cigarette = c("Y", "N")
    )
  )
marijuana_ct_flip

marijuana_data <- 
  tibble(
    Alcohol = 
      as.factor(
        rep(
          c("Y", "N")
        ,times = 
          c(
            sum(marijuana_ct[,,"Y"])
            ,sum(marijuana_ct[,,"N"])
          )
        )
      )
    ,Cigarette =
      as.factor(
        rep(
          rep(c("Y", "N"), times = dim(marijuana_ct)[3])
          ,times =
            c(
              sum(marijuana_ct["Y",,"Y"])
              ,sum(marijuana_ct["N",,"Y"])
              ,sum(marijuana_ct["Y",,"N"])
              ,sum(marijuana_ct["N",,"N"])
            )
        )
      )
    ,Marijuana =
      as.numeric(
        rep(
          rep(c(1, 0), times = dim(marijuana_ct)[3]*dim(marijuana_ct)[1])
          ,times =
            c(
              sum(marijuana_ct["Y","Y","Y"])
              ,sum(marijuana_ct["Y","N","Y"])
              ,sum(marijuana_ct["N","Y","Y"])
              ,sum(marijuana_ct["N","N","Y"])
              ,sum(marijuana_ct["Y","Y","N"])
              ,sum(marijuana_ct["Y","N","N"])
              ,sum(marijuana_ct["N","Y","N"])
              ,sum(marijuana_ct["N","N","N"])
            )
        )
      )
  )
marijuana_summary <- 
  marijuana_data %>% 
  group_by(Alcohol, Cigarette) %>% 
  summarize(
    Yes = sum(Marijuana)
    ,No = length(Marijuana) - Yes
    ,Total = Yes + No
  )
marijuana_summary

marijuana_plot_data <- 
  marijuana_summary %>%
  ungroup %>% 
  mutate(
    `Alcohol Use` = if_else(Alcohol == "Y", "Yes", "No")
    ,`Cigarette Use` = if_else(Cigarette == "Y", "Yes", "No")
    ,Yes = Yes/Total
    ,No = 1 - Yes
  ) %>% 
 dplyr::select(
    -c(
      Alcohol
      ,Cigarette
      ,Total
    )
  ) %>%
  gather(
    key = `Marijuana Use`
    ,value = `Sample Proportion`
    ,-`Alcohol Use`
    ,-`Cigarette Use`
  )

marijuana_plot_data %>% 
  ggplot(
    aes(
      x = `Cigarette Use`
      ,y = `Sample Proportion`
      ,fill = `Marijuana Use`
    )
  ) +
  geom_col(alpha = 0.75) +
  facet_wrap(
    ~ `Alcohol Use`
    ,nrow = 1
  ) +
  scale_fill_brewer(
    type = "qual"
    ,name = "Marijuana Use:"
    ,palette = "Set2"
  ) +
  labs(
    title = "Proportion of Persons with Marijuana Use by Cigarette Use"
    ,subtitle = "Conditional on Alcohol Use"
  )

marijuana_plot_data %>% 
  ggplot(
    aes(
      x = `Alcohol Use`
      ,y = `Sample Proportion`
      ,fill = `Marijuana Use`
    )
  ) +
  geom_col(alpha = 0.75) +
  facet_wrap(
    ~ `Cigarette Use`
    ,nrow = 1
  ) +
  scale_fill_brewer(
    type = "qual"
    ,name = "Marijuana Use:"
    ,palette = "Set2"
  ) +
  labs(
    title = "Proportion of Persons with Marijuana Use by Alcohol Use"
    ,subtitle = "Conditional on Cigarette Use"
  )
```

### Cochran-Mantel-Haenszel Test
$$
  H_{0}:\text{ Marijuana Use is independent of Cigarette Use, conditional on Alcohol Use}
$$
$$
  H_{1}:\text{ Marijuana Use is dependent on Cigarette Use, conditional on Alcohol Use}
$$
```{r 4.31 CMH}
marijuana_cmh <- mantelhaen.test(marijuana_ct, correct = FALSE)
marijuana_cmh
```
With a p-value of `r round(marijuana_cmh$p.value, 4)`, we reject the null hypothesis, and accept the alternative based on the sample.

$$
  H_{0}:\text{ Marijuana Use is independent of Alcohol Use, conditional on Cigarette Use}
$$
$$
  H_{1}:\text{ Marijuana Use is dependent on Alcohol Use, conditional on Cigarette Use}
$$
```{r 4.31 CMH Flip}
marijuana_cmh_flip <- mantelhaen.test(marijuana_ct_flip, correct = FALSE)
marijuana_cmh_flip
```
With a p-value of `r round(marijuana_cmh_flip$p.value, 4)`, we reject the null hypothesis, and accept the alternative based on the sample.

### Logistic Regression
```{r 4.31 Logit}
logit_marijuana_01 <- 
  marijuana_data %>% 
  glm(
    Marijuana ~ Alcohol + Cigarette + Alcohol:Cigarette
    ,family = binomial
    ,data = .
  )
summary(logit_marijuana_01)

logit_marijuana_02 <-
  marijuana_data %>% 
  glm(
    Marijuana ~ Alcohol + Cigarette
    ,family = binomial
    ,data = .
  )
summary(logit_marijuana_02)

anova(logit_marijuana_02, logit_marijuana_01, test = "Chisq")
```
By doing a Likelihood Ratio Test between the model with an interaction term, and the model without, a p-value of `r round(anova(logit_marijuana_02, logit_marijuana_01, test = "Chisq")$"Pr(>Chi)"[2], 4)` is found. This fails to reject the null hypothesis of explaining equal amounts of variance, based on the sample. As such we can use the more parsimonious model.

Both the slope parameters are significant at $\alpha=0.05$, based on the sample. A person with Alcohol Use on average has an increased log likelihood in favour of Marijuana Use of `r round(coefficients(logit_marijuana_02)[2], 4)` vs. a person without, based on the sample. A person with Cigarette Use on average has an increased log likelihood in favour of Marijuana Use of `r round(coefficients(logit_marijuana_02)[3], 4)` vs. a person without, based on the sample.


## 5.17
```{r 5.17}
death_ct <- 
  array(
    data =
      c(
         53, 11, 414, 37
        ,0, 4, 16, 139
      )
    ,dim = c(2, 2, 2)
    ,dimnames = list(
      Defendant = c("W", "B")
      ,Death = c("Y", "N")
      ,Victim = c("W", "B")
    )
  )
death_ct

death_data <- 
  tibble(
    Victim = 
      as.factor(
        rep(
          c("W", "B")
        ,times = 
          c(
            sum(death_ct[,,"W"])
            ,sum(death_ct[,,"B"])
          )
        )
      )
    ,Defendant =
      as.factor(
        rep(
          rep(c("W", "B"), times = dim(death_ct)[3])
          ,times =
            c(
              sum(death_ct["W",,"W"])
              ,sum(death_ct["B",,"W"])
              ,sum(death_ct["W",,"B"])
              ,sum(death_ct["B",,"B"])
            )
        )
      )
    ,Death =
      as.numeric(
        rep(
          rep(c(1, 0), times = dim(death_ct)[3]*dim(death_ct)[1])
          ,times =
            c(
              sum(death_ct["W","Y","W"])
              ,sum(death_ct["W","N","W"])
              ,sum(death_ct["B","Y","W"])
              ,sum(death_ct["B","N","W"])
              ,sum(death_ct["W","Y","B"])
              ,sum(death_ct["W","N","B"])
              ,sum(death_ct["B","Y","B"])
              ,sum(death_ct["B","N","B"])
            )
        )
      )
  )
death_summary <- 
  death_data %>% 
  group_by(Victim, Defendant) %>% 
  summarize(
    Yes = sum(Death)
    ,No = length(Death) - Yes
    ,Total = Yes + No
  )
death_summary

death_plot_data <- 
  death_summary %>%
  ungroup %>% 
  mutate(
    Victim = if_else(Victim == "W", "White", "Black")
    ,Defendant = if_else(Defendant == "W", "White", "Black")
    ,Yes = Yes/Total
    ,No = 1 - Yes
  ) %>% 
 dplyr::select(
    -Total
  ) %>%
  gather(
    key = `Death Penalty`
    ,value = `Sample Proportion`
    ,-Victim
    ,-Defendant
  )

death_plot_data %>% 
  ggplot(
    aes(
      x = Defendant
      ,y = `Sample Proportion`
      ,fill = `Death Penalty`
    )
  ) +
  geom_col(alpha = 0.75) +
  facet_wrap(
    ~ Victim
    ,nrow = 1
  ) +
  scale_fill_brewer(
    type = "qual"
    ,name = "Death Penalty:"
    ,palette = "Set2"
  ) +
  labs(
    title = "Proportion of Defendants who Received the Death Penalty by Defendant's Race"
    ,subtitle = "Conditional on Victim's Race"
  )

death_plot_data %>% 
  ggplot(
    aes(
      x = Victim
      ,y = `Sample Proportion`
      ,fill = `Death Penalty`
    )
  ) +
  geom_col(alpha = 0.75) +
  facet_wrap(
    ~ Defendant
    ,nrow = 1
  ) +
  scale_fill_brewer(
    type = "qual"
    ,name = "Death Penalty:"
    ,palette = "Set2"
  ) +
  labs(
    title = "Proportion of Defendants who Received the Death Penalty by Victims's Race"
    ,subtitle = "Conditional on Defendants's Race"
  )
```

### a.
```{r 5.17a}
logit_death <- 
  death_data %>% 
  glm(
    Death ~ Victim + Defendant
    ,family = binomial
    ,data = .
  )
summary(logit_death)

death_summary %<>% 
  cbind(
    phat = 
      predict(
        object = logit_death
        ,newdata = death_summary %>% dplyr::select(Victim, Defendant)
        ,type = "response"
      ) %>% t
  ) %>% 
  mutate(
    Expected_Yes = Total*phat
    ,Expected_No = Total*(1-phat)
  )
death_summary

death_chisq_stat <-
  death_summary %>% 
  ungroup %>% 
  mutate(
    ChiSq_Yes = (Yes - Expected_Yes)^(2) / Expected_Yes
    ,ChiSq_No = (No - Expected_No)^(2) / Expected_No
  ) %>% 
  dplyr::select(
    ChiSq_Yes
    ,ChiSq_No
  ) %>% 
  sum
death_chisq_stat
death_chisq_p_value <- 
  pchisq(
    q = death_chisq_stat
    ,df = prod(dim(death_ct)-1)
    ,lower.tail = FALSE
  )
death_chisq_p_value
```
Based on the sample, and the logistic regression found above, doing a $\chi^{2}$ Goodness of Fit Test yields a p-value of `r round(death_chisq_p_value, 4)`. Therefore we fail to reject the null hypothesis that the model fits the data and assume it to be true. A $G^{2}$ Goodness of Fit Test can't be used as no White Defendants with Black Victims were given the death penalty.

### c.
Both the slope parameters are significant at $\alpha=0.05$, based on the sample. A Defendant with a White Victim on average has an increased log likelihood in favour of getting the death penalty of `r round(coefficients(logit_death)[2], 4)` vs. a Black Victim, based on the sample. A White Defendant on average has an increased log likelihood in favour of getting the death penalty of `r round(coefficients(logit_death)[3], 4)` vs. a Black Defendant, based on the sample.


## 5.18
```{r 5.18}
lung_cancer_ct <- 
  array(
    data =
      c(
         126, 35, 100, 61
        ,908, 497, 688, 807
        ,913, 336, 747, 598
        ,235, 58, 172, 121
        ,402, 121, 308, 215
        ,182, 72, 156, 98
        ,60, 11, 99, 43
        ,104, 21, 89, 36
      )
    ,dim = c(2, 2, 8)
    ,dimnames = list(
      Smoking = c("Y", "N")
      ,Lung = c("Y", "N")
      ,City = 
        c(
          "Beij"
          ,"Shan"
          ,"Shen"
          ,"Nanj"
          ,"Harb"
          ,"Zhen"
          ,"Taiy"
          ,"Nanc"
        )
    )
  )
lung_cancer_ct

lung_cancer_data <- flatr::flatten_ct(lung_cancer_ct)

lung_cancer_summary <- 
  lung_cancer_data %>% 
  group_by(City, Smoking) %>% 
  mutate(Lung_Num = if_else(Lung == "Y", 1, 0 )) %>% 
  summarize(
    Yes = sum(Lung_Num)
    ,No = length(Lung_Num) - Yes
    ,Total = Yes + No
  )
lung_cancer_summary

lung_cancer_plot_data <- 
  lung_cancer_summary %>%
  ungroup %>% 
  mutate(
    Smoking = if_else(Smoking == "Y", "Yes", "No")
    ,Yes = Yes/Total
    ,No = 1 - Yes
  ) %>% 
 dplyr::select(
    -Total
  ) %>%
  gather(
    key = `Lung Cancer`
    ,value = `Sample Proportion`
    ,-City
    ,-Smoking
  )

lung_cancer_plot_data %>% 
  ggplot(
    aes(
      x = Smoking
      ,y = `Sample Proportion`
      ,fill = `Lung Cancer`
    )
  ) +
  geom_col(alpha = 0.75) +
  facet_wrap(
    ~ City
    ,nrow = 1
  ) +
  scale_fill_brewer(
    type = "qual"
    ,name = "Has Lung Cancer:"
    ,palette = "Set2"
  ) +
  labs(
    title = "Proportion of Persons with Lung Cancer by whether or not they Smoke"
    ,subtitle = "Conditional on City"
  )

lung_cancer_plot_data %>% 
  ggplot(
    aes(
      x = City
      ,y = `Sample Proportion`
      ,fill = `Lung Cancer`
    )
  ) +
  geom_col(alpha = 0.75) +
  facet_wrap(
    ~ Smoking
    ,nrow = 1
  ) +
  scale_fill_brewer(
    type = "qual"
    ,name = "Has Lung Cancer:"
    ,palette = "Set2"
  ) +
  labs(
    title = "Proportion of Persons with Lung Cancer by City"
    ,subtitle = "Conditional on whether or not they Smoke"
  ) +
  theme(
    axis.text.x = element_text(
      angle = 45
      ,hjust = 0.5
      ,vjust = 0.5
    )
  )
```

### a.
```{r 5.18a}
logit_lung_cancer <- 
  lung_cancer_data %>% 
  glm(
    Lung ~ City + Smoking
    ,family = binomial
    ,data = .
  )
summary(logit_lung_cancer)
```
A Smoker on average has an increased log likelihood in favour of getting Lung Cancer of `r round(coefficients(logit_lung_cancer)[9], 4)` vs. a non-smoker, based on the sample.

### b.
```{r 5.18b}
lung_cancer_summary %<>% 
  cbind(
    phat = 
      predict(
        object = logit_lung_cancer
        ,newdata = lung_cancer_summary %>% dplyr::select(City, Smoking)
        ,type = "response"
      ) %>% t
  ) %>% 
  mutate(
    Expected_Yes = Total*phat
    ,Expected_No = Total*(1-phat)
  )
lung_cancer_summary

chisq_gof <- flatr::goodness_of_fit(model = logit_lung_cancer, response = "Lung", type = "Chisq")
chisq_gof

gsq_gof <- flatr::goodness_of_fit(model = logit_lung_cancer, response = "Lung", type = "Gsq")
gsq_gof
```
Based on the sample, and the logistic regression found above, doing a $\chi^{2}$ Goodness of Fit Test yields a p-value of `r round(chisq_gof[[2]], 4)`. Therefore we fail to reject the null hypothesis that the model fits the data and assume it to be true.

Based on the sample, and the logistic regression found above, doing a $G^{2}$ Goodness of Fit Test yields a p-value of `r round(gsq_gof[[2]], 4)`. Therefore we fail to reject the null hypothesis that the model fits the data and assume it to be true.