---
title: "Week 03 Questions"
author: "Scott Graham"
date: "September 29, 2017"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse, warn.conflicts = FALSE, quietly = TRUE)
library(magrittr, warn.conflicts = FALSE, quietly = TRUE)
```
\[
  \newcommand{\Prob}{\operatorname{P}}
  \newcommand{\E}{\operatorname{E}}
  \newcommand{\Var}{\operatorname{Var}}
  \newcommand{\Cov}{\operatorname{Cov}}
  \newcommand{\se}{\operatorname{se}}
  \newcommand{\re}{\operatorname{re}}
  \newcommand{\ybar}{{\overline{Y}}}
  \newcommand{\phat}{{\hat{p}}}
  \newcommand{\that}{{\hat{T}}}
  \newcommand{\med}{{\tilde{Y}}}
\]

## 2.29
Let $\alpha=0.05$, and:
$$
  H_{0}: OR = \frac{Odds_{Drug}}{Odds_{Control}} \leq 1
$$
$$
  H_{1}: OR \frac{Odds_{Drug}}{Odds_{Control}} > 1
$$
```{r 2.29}
results_mat <- matrix(nrow = 2, ncol = 2, c(7, 0, 8, 15))
rownames(results_mat) <- c("Drug", "Control")
colnames(results_mat) <- c("Normalized", "Not Normalized")
results_mat
fisher.test(results_mat, alternative = "greater")
```
Since our p-value (0.003161) is $<\alpha$, we reject our null hypothesis that the odd's ratios for the drug and the control $=1$, and assume that the odds ratio is greater than 1.


## 2.31
### a.
Let $\alpha=0.05$, and:
$$
  H_{0}: OR = \frac{Odds_{Surgery}}{Odds_{Radiation}} = 1
$$
$$
  H_{1}: OR \frac{Odds_{Surgery}}{Odds_{Radiation}} \neq 1
$$
```{r 2.31a}
results_mat <- matrix(nrow = 2, ncol = 2, c(21, 15, 2, 3))
rownames(results_mat) <- c("Surgey", "Radiation")
colnames(results_mat) <- c(" Cancer Controlled", "Cancer Not Controlled")
results_mat
fisher.test(results_mat, alternative = "two.sided")
1 - dhyper(
  x = results_mat[1, 1] - 1
  ,m = sum(results_mat[, 1])
  ,n = sum(results_mat[, 2])
  ,k = sum(results_mat[1, ])
  )
```
Because our p-value (0.6384) $>\alpha$, we fail to reject our null hypothesis, and assume the odds of controlling cancer are the same if you have surgery or radiation based on the sample.
$$
  P-Value = 
  \sum_{i=18}^{23}p_{i} \leq
  \Prob\left( n_{11} = 21 \right)
$$
$$
  P-Value =
  \Prob\left( n_{11} = a \right) >
  \Prob\left( n_{11} = 21 \right) \implies
  1 - \Prob\left( n_{11} = 20 \right)
$$

### b.
```{r 2.31b}
1 -
  phyper(
    q = results_mat[1, 1]
    ,m = sum(results_mat[, 1])
    ,n = sum(results_mat[, 2])
    ,k = sum(results_mat[1, ])) + 0.5*
  dhyper(
    x = results_mat[1, 1]
    ,m = sum(results_mat[, 1])
    ,n = sum(results_mat[, 2])
    ,k = sum(results_mat[1, ]))
```
Because our p-value (0.2431) $>\alpha$, we fail to reject our null hypothesis, and assume the odds of controlling cancer are the same if you have surgery or radiation based on the sample.


## 2.39
### a.
True

### b.
True

### c.
False

### d.
True

### e.
False


## 3.3
### a.
$$
  \E\left[ \Prob\left( Y = 1 \middle| X = x \right) \right] =
  0.00255+0.00109x,
  x = \{0, 0.5, 1.5, 4.0, 7.0\}
$$
That is, on average with no alcohol consumption, there is a $0.255\%$ probability of a sex organ malformation. And on average, for each unit increase in $x$ (alcohol consumption), there is a $0.108\%$ increase in the probability of sex organ malformation, based on the sample. 

### b.
$$
  \Prob\left( Y = 1 \middle| X = (0, 7.0) \right)=
  0.00255+0.00109x =
  (0.00255+0.00109(0)), 0.00255+0.00109(7.0)) =
  (0.00255, 0.01018) \implies
$$
$$
  RR =
  \frac{\Prob\left( Y = 1 \middle| X = 7.0 \right)}{\Prob\left( Y = 1 \middle| X = 0 \right)} =
  \frac{0.01018}{0.00255} =
  3.992157
$$


## 3.9
```{r 3.9}
income.credit <- read_csv("http://grahamst.at/projects/STAT_541/STAT_541_Assignments/Week_03_Data_3.09.csv")
income.credit
income.credit.glm <- glm(Card ~ Inc, data = income.credit, family = "binomial")
summary(income.credit.glm)
```
### a.
$$
  \ln \left( \frac{\Prob\left( Y = 1 \middle| X = x \right)}{1 - \Prob\left( Y = 1 \middle| X = x \right)} \right) =
  -3.5561 + 0.0532x \implies
$$
$$
  \Prob\left( Y = 1 \middle| X = x \right) = 
  \frac{e^{-3.5561 + 0.0532x}}{1+e^{-3.5561 + 0.0532x}}
$$

### b.
On average, as the annual income in Lira increase, the probability of owning at least one travel credit card increases as well, based on the sample.

### c.
$$
  \ln \left( \frac{0.5}{1 - 0.5} \right) =
  -3.5561 + 0.0532x \implies
  0 =
  -3.5561 + 0.0532x \implies
  x =
  66.84398 \approx
  66.86
$$
When our LHS = 0, we can solve for $x$, which gives $x\approx 66.86$ million Lira. That is, on average there is a $50\%$ chance of possessing at least one travel credit card  when one's annual income is 66.86 million Lira, based on the sample.