---
title: "Week 10 Questions"
author: "Scott Graham"
date: "Novemeber 17, 2017"
output:
  html_document:
    theme: lumen
    toc: true
    toc_depth: 2
    toc_float: true
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(nnet, warn.conflicts = FALSE, quietly = TRUE)
library(tidyverse, warn.conflicts = FALSE, quietly = TRUE)
library(magrittr, warn.conflicts = FALSE, quietly = TRUE)
library(flatr, warn.conflicts = FALSE, quietly = TRUE)
theme_minimal2 <- theme_minimal() %>%  theme_set()
theme_minimal2 <-
  theme_update(
    panel.border = element_rect(
      linetype = "solid"
      ,colour = "grey92"
      ,fill = NA
    )
    ,strip.background = element_rect(
      linetype = "solid"
      ,colour = "grey92"
      ,fill = NA
    )
  )
```
\[
  \newcommand{\Prob}{\operatorname{P}}
  \newcommand{\E}{\operatorname{E}}
  \newcommand{\Var}{\operatorname{Var}}
  \newcommand{\Cov}{\operatorname{Cov}}
  \newcommand{\se}{\operatorname{se}}
  \newcommand{\re}{\operatorname{re}}
  \newcommand{\ybar}{{\overline{Y}}}
  \newcommand{\phat}{{\hat{p}}}
  \newcommand{\that}{{\hat{T}}}
  \newcommand{\med}{{\tilde{Y}}}
\]

## 6.01
### a.
$$
  \ln\left( \frac{\hat{\pi}_{R}}{\hat{\pi}_{D}} \right) =
   \ln\left( \frac{\hat{\pi}_{R}}{\hat{\pi}_{I}} \right) -  \ln\left( \frac{\hat{\pi}_{D}}{\hat{\pi}_{I}} \right) =
   -2.3 + 0.5x
$$

For every \$10,000 increase in annual income, the estimated log odds in favour of being a Republican vs. a Democrat increases by 0.5.

### b.
$$
    -2.3 + 0.5x > 0 \implies
    0.5x > 2.3 \implies
    x > 4.6
$$

### c.
$$
  \hat{\pi}_{I} = 
  \frac{e^{0}}{e^{0} + e^{3.3 ??? 0.2x} + e^{1.0 + 0.3x}} =
  \frac{1}{1 + e^{3.3 ??? 0.2x} + e^{1.0 + 0.3x}}
$$


## 6.03
```{r 6.03}
gator_ct <- 
  array(
    data =
      c(
        23, 7, 4, 0, 2, 1, 2, 3, 8, 5
        ,5, 13, 11, 8, 1, 6, 0, 1, 3, 0
        ,5, 8, 11, 7, 2, 6, 1, 3, 5, 5
        ,16, 17, 19, 1, 1, 0, 2, 1, 3, 3
      )
    ,dim = c(2, 5, 4)
    ,dimnames = list(
      Size = c("<=2.3", ">2.3")
      ,Food = c(
        "Fish"
        ,"Invertebrate"
        ,"Reptile"
        ,"Bird"
        ,"People/Other"
      )
      ,Lake = 
        c(
          "Hancock"
          ,"Oklawaha"
          ,"Trafford"
          ,"George"
        )
    )
  )
gator_ct
```
### a.
```{r 6.03a}
gator_logit <- 
  gator_ct %>% 
  flatten_ct() %>% 
  multinom(
    Food ~ Size + Lake
    ,data = .
  )
summary(gator_logit)
```
$$
  \ln\left( \frac{\pi_{i}}{\pi_{Bird}} \right) =
  \frac{e^{\alpha_{i} + \beta_{1i}Size + \beta_{2i}LakeHancock + \beta_{3i}LakeOklawaha + \beta_{4i}LakeTrafford}}{1 + \sum_{\forall i}e^{\alpha_{i} + \beta_{1i}Size + \beta_{2i}LakeHancock + \beta_{3i}LakeOklawaha + \beta_{4i}LakeTrafford}}
$$
With the $\alpha$s and $\beta$s from above.

### b.
```{r 6.03b}
predict(
  object = gator_logit
  ,newdata = 
    data.frame(
      Lake = "Oklawaha"
      ,Size = c("<=2.3", ">2.3")
    ),
  type = "probs"
)
```


## 6.05
### a.
As $x_{1}$ increases, you become more satisfied $(\beta_{1}<0)$.

As $x_{2}$ increases, you become less satisfied $(\beta_{2}>0)$.

As $x_{3}$ increases, you become less satisfied $(\beta_{3}>0)$.

### b.
$$
  x_{1} = 4,
  x_{2} = 1, 
  x_{3} = 1 \implies
  \operatorname{Logit}\left[ \hat{\Prob} \left( Y \leq j \right) \right] =
  \hat{\alpha}_{j} - 2.16 + 0.60 + 1.19 =
  \hat{\alpha}_{j} - 0.37
$$


## 6.07
### a.
$\beta$ is held constant in a Cumulative Logit, and $\alpha$ varies.

### b.
(I think this is SAS)

For every unit inrease in income, the estimated log odds in favour of an unhappy marriage decrease to that of a happy marriage by 0.1117.

### c.
Test Statistic $=08976$, P-Value $=0.3434$. We then fail to reject the null hypothesis that income doesn't have an effect.

### d.
Yes as it has no statistically significant deviance.

### e.
$$
  \Prob(\text{Very Happy Marriage}) = 
  1 - \Prob(\text{Pretty Happy Marriage}) =
  1- \frac{e^{-0.2378-0.1117(2)}}{1+e^{-0.2378-0.1117(2)}} = 0.6133
$$


## 6.09
### a.
Because there are $5$ categories, we only need to estimate $5-1=4$ categories.

### b.
Most Liberal: None as it has the highest $\beta=0$
Most Conservative: Protestant as it has the smallest $\beta=-1.27$

### c.
$$
  \Prob\left( \text{Liberal} \middle | \text{Protestant} \right) =
  \frac{e^{-1.03-1.27}}{1 + e^{-1.03-1.27}} = 
  0.0911
$$
$$
  \Prob\left( \text{Liberal} \middle | \text{None} \right) =
  \frac{e^{-1.03}}{1 + e^{-1.03}} = 
  0.2631
$$

### d.
$$
  e^{-1.27} =
  0.2808
$$
The estimated odds of being more liberal for a protestant rather than none is 0.2808 times higher.
$$
  e^{-1.27-(-1.22)} =
  0.9512
$$
The estimated odds of being more liberal for a protestant rather than catholic is 0.9512 times higher.