---
title: "Week 01 Questions"
author: "Scott Graham"
date: "September 15, 2017"
output:
  html_document:
    theme: lumen
    toc: true
    toc_depth: 2
    toc_float: true
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
\[`
  \newcommand{\Prob}{\operatorname{P}}
  \newcommand{\E}{\operatorname{E}}
  \newcommand{\Var}{\operatorname{Var}}
  \newcommand{\Cov}{\operatorname{Cov}}
  \newcommand{\se}{\operatorname{se}}
  \newcommand{\re}{\operatorname{re}}
  \newcommand{\ybar}{{\overline{Y}}}
  \newcommand{\phat}{{\hat{p}}}
  \newcommand{\that}{{\hat{T}}}
  \newcommand{\med}{{\tilde{Y}}}
\]

## 1.01

### a.
The explanatory variables are Gender (male, female), and Mother's education (high school, college). The response variable is Attitude towards gun control (favour, oppose).

### b.
The explanatory variables are Blood pressure, and Cholesterol level. The response variable is Heart Disease (yes, no).

### c.
The explanatory variables are Race (white, nonwhite), Religion (Catholic, Jewish, Protestant), and Annual income. The response variable is Vote for president (Democrat, Republican, Other).

### d.
The explanatory variable is Marital status (married, single, divorced, widowed). The response variable is Quality of life (excellent, good, fair, poor).


## 1.03

### a.
Let $X=$ the number of correct answers by the student on the exam. Let $p=\frac{1}{4}$ be the probability of guessing the correct answer for a given question. Let $n=100$ be the number of questions on the exam. Then:
$$
  X \sim Binom\left( n = 100, p = \frac{1}{4} \right)
$$

### b.
Assuming a Type I error rate of $5\%$:
$$
  H_{0}: p = 0.25
$$
$$
  H_{1}: p > 0.25
$$
$$
  \Prob \left( X \geq 50 \right) =
  \sum_{x=50}^{100} {{100}\choose{x}} 0.25^{x} (1-0.25)^{100-x}
$$
or equivalently in R:
```{r 1.3b}
  pbinom(50 - 1, 100, 0.25, lower.tail = FALSE)
```
which is $<5\%$. As such we reject the null hypothesis that the student was merely guessing for each question, and assume that their probability of getting the correct answer is $>0.25$ based on the sample.


## 1.05
We need to find $\phat$ s.t. we maximize $\Prob\left(Y = 0 \middle| n = 2 \right)$ subject to $p\in[0,1]$
$$
  \Prob\left(Y = 0 \middle| n = 2 \right) =
  {{2}\choose{0}} p^{0} (1-p)^{2-0} =
  (1-p)^{2} \implies
$$
$$
  \frac{d}{dp} (1-p)^{2} =
  -2(1-p) \implies
  0 = 
  -2(1-p) \implies
  p = 1
$$
$p=0$ is also an endpoint, so we can check that as well. For $p=1$, we have a likelihood of 0, and for $p=0$, we have a likelihood of 1.
```{r 1.5}
# p = 0
dbinom(x = 0, size = 2, prob = 0)
# p = 1
dbinom(x = 0, size = 2, prob = 1)
```
Since our function is monotonic over this domain, $p = 0$ is our MLE.

This may not be reasonable however, as this would require the coin to have both sides as tails. As well, we have a small sample size of $n=2$, which makes it hard to draw a reasonable conclusion. If we use the bayesian estimator as given in the text, $p=\frac{1}{4}$, which may be a more reasonable estimator for $p$ based on our limited data.


## 1.07

### a.
Let $X$ be the bullet firing (a somewhat morbid success...), $n=6$, $p=\frac{1}{6}$:
$$
  \Prob \left( X = 0 \middle| n = 6, p = \frac{1}{6} \right) =
  {{6}\choose{0}} \left( \frac{1}{6} \right)^{0} \left( 1 - \frac{1}{6} \right)^{6-0} =
  \left( \frac{5}{6} \right)^{6} =
  0.3349
$$

### b.
We have the case where we have $y-1$ number of non-fires, and then 1 fire. Since the order of the non-fires doesn't matter, we don't need a binomial term in our equation, as it would = 1. As well, each spin is independent, which allows us to multiply our probabilities together.
$$
  \Prob \left( X = 0 \middle| n = y - 1 \right) \times \Prob \left( X = 1 \middle| n = 1 \right) =
  (1-p)^{y-1}p \stackrel{{p=\frac{1}{6}}}{=}
  \left( \frac{5}{6} \right)^{y-1} \left( \frac{1}{6} \right)
$$


## 1.11

$$
  \phat = 0.86, n=1158
$$
$$
  \se \left( \phat \right) = 
  \sqrt{\frac{0.86(1-0.86)}{1158}} =
  0.0102
$$
95% CI for $p$:
$$
  0.86 \mp 1.96(0.0102) =
  [0.84, 0.88]
$$

That is, we can say with 95% confidence the true proportion of american adults that believe in heaven is somewhere between 84% and 88% based on our sample.


## 1.13

### a.
$$
  \ell = 
  \prod_{i=1}^{25} \Prob \left( Y_{i} = 0 \right) =
  \prod_{i=1}^{25} {{25}\choose{0}} p^{y_{i}} (1-p)^{25-y_{i}} =
$$
Note: We can ignore the binomial term at the start, as it doesn't contain $p$.
$$
  \ell =
  p^{\sum_{i=1}^{25} y_{i}} (1-p)^{25-\sum_{i=1}^{25} y_{i}}
$$
Since $y_{i}=0,\forall 1=1,2,\dots,25$:
$$
  \ell = (1-p)^{25}
$$

Under $H_{0}$, we assume $p=0.5$, our likelihood function ($\ell_{0}$) is maximized at $p=0.5$.
$$
  \implies \ell_{0} = (0.5)^{25}
$$

### b.
$$
  \ell_{1} = (1-p)^{25} \implies
  \frac{d}{dp}(1-p)^{25} = -25(1-p)^{24} \implies
  0 = -25(1-p)^{24} \implies
  p = 1
$$
We also get $p=0$ due to it being our other endpoint in the domain $p\in[0,1]$.
$$
  \ell_{1}(0) = 
  (1-0)^{25} = 
  1
$$
$$
  \ell_{1}(1) = 
  (1-1)^{25} = 
  0
$$

So our MLE is 0, with a likelihood of 1.

### c.
$$
  {\chi^{2}}_{calc} =
  -2\ln \left( \frac{(1-0.5)^{25}}{(1-0)^{25}} \right) =
  -2 \ln \left( 0.5^{25} \right) =
  34.65736 \approx
  34.7
$$
```{r 1.13c}
  pchisq(q = -2*log(0.5^25), df = 1, lower.tail = FALSE)
```

So we have a p-value of $3.93 \times 10^{-9}$, and can then reject our null hypothesis at that level.

### d.
$$
  {\chi^{2}}_{calc} =
  -2\ln \left( \frac{(1-0.074)^{25}}{(1-0)^{25}} \right) =
  3.844052 \approx
  3.84
$$
```{r 1.13d}
  pchisq(q = -2*log((1-0.074)^25), df = 1, lower.tail = FALSE)
```
So our p-value is $\approx 0.05$ meaning it is the upper bound for the 95% likelihood-ratio confidence interval.