---
title: "Analysis of EA NHL Results"
author: "Scott Graham"
date: "December 06, 2017"
output:
  ioslides_presentation:
    smaller: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

set.seed(5609)

library(knitr, warn.conflicts = FALSE, quietly = TRUE)
library(kableExtra, warn.conflicts = FALSE, quietly = TRUE)
library(tidyverse, warn.conflicts = FALSE, quietly = TRUE)
library(ggfortify, warn.conflicts = FALSE, quietly = TRUE)
library(magrittr, warn.conflicts = FALSE, quietly = TRUE)
library(flatr, warn.conflicts = FALSE, quietly = TRUE)
library(lmtest, warn.conflicts = FALSE, quietly = TRUE)
library(glmnet, warn.conflicts = FALSE, quietly = TRUE)

theme_minimal2 <- theme_minimal() %>%  theme_set()
theme_minimal2 <-
  theme_update(
    panel.border = element_rect(
      linetype = "solid"
      ,colour = "grey92"
      ,fill = NA
    )
    ,strip.background = element_rect(
      linetype = "solid"
      ,colour = "grey92"
      ,fill = NA
    )
  )

hockey_data <- 
  "http://grahamst.at/projects/STAT_541/Final_Project/data/hockey_stats.csv" %>% 
  read_csv() %>% 
  rename(`Day of the Week` = `Day of Week`) %>% 
  mutate(
    `P1 Win` = as.factor(`P1 Win`)
    ,`P2 - P1 Goals` = P2 - P1
    ,`Shot Differential` = `P2 Shots` - `P1 Shots`
    ,`Hit Differential` = `P2 Hits` - `P1 Hits`
    ,`Day of the Week` = as.factor(`Day of the Week`)
    ,`Scored 1st Boolean (P2)` = if_else(`Scored 1st` == "P2", 1, 0)
    ,Shots = `P1 Shots` + `P2 Shots`
    ,Hits = `P1 Hits` + `P2 Hits`
    ,ToA = `P1 ToA` + `P2 ToA`
    ,Passing = `P1 Passing` + `P2 Passing`
    ,PM = `P1 PM` + `P2 PM`
    ,Powerplays = `P1 Powerplays` + `P2 Powerplays`
    ,`Penalty Shots` = `P1 Penalty Shots` + `P2 Penalty Shots`
    ,Faceoffs = `P1 Faceoffs` + `P2 Faceoffs`
    ,`Offensive Faceoffs` = `P1 Offensive Faceoffs` + `P2 Offensive Faceoffs`
    ,Breakaways = `P1 Breakaways` + `P2 Breakaways`
  )

day_of_week_conv <- 
  c(
    "0" = "Sunday"
    ,"1" = "Monday"
    ,"2" = "Tuesday"
    ,"3" = "Wednesday"
    ,"4" = "Thursday"
    ,"5" = "Friday"
    ,"6" = "Saturday"
  )

```
\[
  \newcommand{\Prob}{\operatorname{P}}
  \newcommand{\E}{\operatorname{E}}
  \newcommand{\Var}{\operatorname{Var}}
  \newcommand{\Cov}{\operatorname{Cov}}
  \newcommand{\se}{\operatorname{se}}
  \newcommand{\re}{\operatorname{re}}
  \newcommand{\ybar}{{\overline{Y}}}
  \newcommand{\phat}{{\hat{p}}}
  \newcommand{\that}{{\hat{T}}}
  \newcommand{\med}{{\tilde{Y}}}
  \newcommand{\Logit}{{\operatorname{Logit}}}
  \]

# Introduction
## Data
- 217 Games Played
- EA NHL 12, 13, and 17
- Tim $\rightarrow$ P1
- Randy $\rightarrow$ P2


## Explanatory Variables
- Categorical:
    - Home vs. Away
    - Scored First

- Discrete:
    - Shots
    - Hits
    - Penalty Minutes (PM)
    - Penalty Shots
    - Breakaways
    - Powerplays
    - Faceoffs Won
    - Offensive Faceoffs Won

- Continuous:
    - Time of Attack (ToA)
    - Passing \%


## Research Question
What periphery (non-goal related) statistics can be used to predict the outcome of a hockey game?



# Exploratory Analysis
## Game Results
```{r Game Results}
hockey_data %>% 
  mutate(
    `P2` = if_else(`P2 - P1 Goals` > 0, `P2 - P1 Goals`, as.integer(0))
    ,`P1` = if_else(`P2 - P1 Goals` <= 0, `P2 - P1 Goals`, as.integer(0))
  ) %>% 
  select(
    Game
    ,`P1`
    ,`P2`
  ) %>% 
  gather(
    key = Player
    ,value = `Goal Differential`
    ,-Game
  ) %>% 
  filter(`Goal Differential` != 0) %>% 
  ggplot(
    aes(
      x = Game
      ,y = `Goal Differential`
      ,fill = `Player`
    )
  ) +
  geom_vline(
    data = 
      hockey_data %>% 
      group_by(Version) %>% 
      summarize(Game = min(Game))
    ,aes(xintercept = Game)
    ,linetype = "dashed"
    ,size = 0.5
  ) +
  geom_col() +
  scale_fill_brewer(
    type = "qual"
    ,palette = "Set2"
  )
```


## Correlation Matrix
```{r Cor Matrix}
hockey_data %>%
  select_if(.predicate = is.numeric) %>%
  select(
    -c(
      Game
      ,Version
      ,`P2 - P1 Goals`
    )
    ,-ends_with("Differential")
  ) %>% 
  cor() %>% 
  as.data.frame() %>%  
  rownames_to_column() %>% 
  as.tibble() %>% 
  gather(
    key = Column
    ,value = Correlation
    ,-rowname
  ) %>% 
  rename(Row = rowname) %>% 
  ggplot(
    aes(
      x = Column
      ,y = Row
      ,fill = Correlation
    )
  ) +
  geom_raster() +
  scale_fill_distiller(
    type = "div"
    ,palette = "RdBu"
    ,limits = c(-1, 1)
  ) +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1)
    ,axis.title.x = element_blank()
    ,axis.title.y = element_blank()
  )
```


## Distribution of Goals
```{r Goal Distribution}
hockey_data %>% 
  select(
    P1
    ,P2
    ,`P2 - P1 Goals`
  ) %>% 
  gather(
    key = "Variable"
    ,value = "Goals"
  ) %>% 
  mutate(
    Variable = 
      factor(
        x = Variable
        ,levels = c("P1", "P2", "P2 - P1 Goals")
      )
  ) %>% 
  ggplot(
    aes(
      x = Goals
      ,colour = Variable
    )
  ) +
  geom_density() +
  facet_wrap(
    facet = ~ Variable
    ,ncol = 1
  ) +
  scale_colour_brewer(
    type = "qual"
    ,name = "Variable"
    ,palette = "Set2"
  )
```


## Goal Distribution by Version
```{r Goal Distribution by Version}
hockey_data %>% 
  select(
    P1
    ,P2
    ,`P2 - P1 Goals`
    ,Version
  ) %>% 
  gather(
    key = "Variable"
    ,value = "Goals"
    ,-Version
  ) %>% 
  mutate(
    Variable = 
      factor(
        x = Variable
        ,levels = c("P1", "P2", "P2 - P1 Goals")
      )
  ) %>% 
  ggplot(
    aes(
      x = Goals
      ,colour = Variable
    )
  ) +
  geom_density() +
  facet_grid(facet = Version ~ Variable) +
  scale_colour_brewer(
    type = "qual"
    ,name = "Variable"
    ,palette = "Set2"
  )
```


## Goals vs. Shots
> "You miss 100% of the shots you don't take."
> - Wayne Gretzky

```{r GvS}
hockey_data_GvS <- 
  hockey_data %>% 
  select(
    Game
    ,`P1 Goals` = P1
    ,`P2 Goals` = P2
    ,`P1 Shots`
    ,`P2 Shots`
  ) %>% 
  gather(
    key = Variable
    ,value = Value
    ,-Game
  ) %>% 
  mutate(
    Player = 
      if_else(
        grepl(pattern = "P1 *", x = .$Variable), "P1", "P2"
      )
    ,Statistic =
      if_else(
        grepl(pattern = "* Goals", x = .$Variable), "Goals", "Shots"
      )
  ) %>% 
  select(-Variable) %>% 
  spread(
    key = Statistic
    ,value = Value
  ) %>% 
  mutate(Player = as.factor(Player))

hockey_data_GvS %>% 
  ggplot(
    aes(
      x = Shots
      ,y = Goals
      ,colour = Player
    )
  ) +
  geom_point(
    data = 
      hockey_data_GvS %>% 
      select(-Player)
    ,colour="grey92"
    ,alpha = 0.75
  ) +
  geom_point() +
  geom_smooth(
    method = "loess"
    ,se = FALSE
  ) +
  geom_smooth(
    method = "lm"
    ,se = FALSE
  )  +
  facet_grid(facet = ~ Player) +
  scale_colour_brewer(
    type = "qual"
    ,name = "Player"
    ,palette = "Set2"
  )
```


## Autocorrelation
Autocorrelation measures the correlation between $\left(X_{t-k}, X_{t}\right), t=k, k+1\dots,n$. It is primarily used in timeseries analysis to measure the tendency for a time series to be mean reverting (negative autocorrelation) or a tendency to go on long runs in one direct or the other (positive autocorrelation). This is primarily used in the analysis of stock returns, but can be applied to any time series.


## Autocorrelation Plot
```{r Autocorrelation Plot}
hockey_data_acf <- 
  hockey_data %>%
  mutate(P1_Win_Bool = if_else(`P1 Win` == "Yes", 1, 0)) %>% 
  select(P1_Win_Bool) %>% 
  acf(plot = FALSE)

data.frame(Lag = hockey_data_acf$lag, Autocorrelation = hockey_data_acf$acf) %>% 
  ggplot(
    aes(
      x = Lag
      ,y = Autocorrelation
    )
  ) +
  geom_hline(
    aes(
      yintercept = 0
    )
    ,linetype = "dotted"
  ) +
  geom_segment(
    aes(
      xend = Lag
      ,yend = 0
    )
  ) +
  geom_smooth(
    linetype = 0
    ,method = "lm"
    ,formula = y ~ 1
  )
```


## Autocorrelation Model
$$
  \operatorname{Logit}\left( \text{P1 Wins} \right) =
  \alpha + \beta_{1}\text{Previous Game's Result}
$$

```{r Autocorrelation Logit}
hockey_acf_logit <- 
  hockey_data %>%
  glm(
    formula = `P1 Win` ~ `P1 Previous Game`
    ,family = binomial
    ,data = .
  )
summary(hockey_acf_logit)
```


## Durbin-Watson Test
$$
  H_{0}: \text{Autocorrelation} = 0
$$
$$
  H_{1}: \text{Autocorrelation} \neq 0
$$
```{r Autocorrelation Test}
lmtest::dwtest(hockey_acf_logit, alternative = "two.sided")
```



# Modeling
## Contingency Table Model
```{r CT}
hockey_ct_summary <- 
  hockey_data %>% 
  select(
    `P1 Win`
    ,`P1 Home`
    ,`Scored 1st`
  ) %>% 
  group_by_all() %>% 
  count() %>% 
  ungroup()

hockey_ct <- 
  array(
    data =
      c(
        hockey_ct_summary %>% 
          filter(
            `P1 Win` == "Yes"
            ,`P1 Home` == "H"
            ,`Scored 1st` == "P1"
          ) %>% 
          select(n) %>% 
          as.numeric()
        ,hockey_ct_summary %>% 
          filter(
            `P1 Win` == "Yes"
            ,`P1 Home` == "H"
            ,`Scored 1st` == "P2"
          ) %>% 
          select(n) %>% 
          as.numeric()
        ,hockey_ct_summary %>% 
          filter(
            `P1 Win` == "No"
            ,`P1 Home` == "H"
            ,`Scored 1st` == "P1"
          ) %>% 
          select(n) %>% 
          as.numeric()
        ,hockey_ct_summary %>% 
          filter(
            `P1 Win` == "No"
            ,`P1 Home` == "H"
            ,`Scored 1st` == "P2"
          ) %>% 
          select(n) %>% 
          as.numeric()
        ,hockey_ct_summary %>% 
          filter(
            `P1 Win` == "Yes"
            ,`P1 Home` == "A"
            ,`Scored 1st` == "P1"
          ) %>% 
          select(n) %>% 
          as.numeric()
        ,hockey_ct_summary %>% 
          filter(
            `P1 Win` == "Yes"
            ,`P1 Home` == "A"
            ,`Scored 1st` == "P2"
          ) %>% 
          select(n) %>% 
          as.numeric()
        ,hockey_ct_summary %>% 
          filter(
            `P1 Win` == "No"
            ,`P1 Home` == "A"
            ,`Scored 1st` == "P1"
          ) %>% 
          select(n) %>% 
          as.numeric()
        ,hockey_ct_summary %>% 
          filter(
            `P1 Win` == "No"
            ,`P1 Home` == "A"
            ,`Scored 1st` == "P2"
          ) %>% 
          select(n) %>% 
          as.numeric()
      )
    ,dim = c(2, 2, 2)
    ,dimnames = 
      list(
        Scored_1st = c("P1", "P2")
        ,Winner = c("P1", "P2")
        ,Home = c("P1", "P2")
      )
  )
hockey_ct
```


## flatten_ct
```{r CT Flatten}
hockey_ct %>% 
  flatten_ct() %>% 
  head() %>% 
  kable(digits = 4, format = "html") %>% 
  kable_styling()

hockey_ct %>% 
  flatten_ct() %>% 
  group_by_all() %>% 
  count() %>% 
  kable(digits = 4, format = "html") %>% 
  kable_styling(bootstrap_options = c("striped", "hover"))

```


## Regression Models and Goodness of Fit Tests
### Logit Model
$$
  \operatorname{Logit}\left( \text{P1 Wins} \right) =
  \alpha + \beta_{1}\text{Scored First} + \beta_{2}\text{Home Team}
$$

### Log Linear Model
$$
  \ln(\mu) =
  \alpha + \beta_{1}\text{Scored First} + \beta_{2}\text{P1 Wins} + \beta_{3}\text{Home Team}
$$


## Regression Models and Goodness of Fit Tests Outputs
```{r CT Model}
hockey_ct_logit <- 
  hockey_ct %>% 
  flatten_ct() %>% 
  glm(
    Winner ~ Scored_1st + Home
    ,family = binomial
    ,data = .
  )
coef(hockey_ct_logit)
goodness_of_fit(model = hockey_ct_logit, type = "Chisq")
goodness_of_fit(model = hockey_ct_logit, type = "Gsq")
```

## Log Linear Models and Goodness of Fit Tests Outputs
```{r CT Loglin Model}
hockey_ct_loglin <- 
  hockey_ct %>% 
  flatten_ct() %>% 
  group_by_all() %>% 
  count() %>% 
  ungroup() %>% 
  glm(
    n ~ (Scored_1st + Winner + Home)^2
    ,family = poisson
    ,data = .
  )
coef(hockey_ct_loglin)
goodness_of_fit_loglin(model = hockey_ct_loglin, type = "Chisq")
goodness_of_fit_loglin(model = hockey_ct_loglin, type = "Gsq")
```


## Full Model
```{r Full Model 02}
hockey_full_logit_02<- 
  glm(
    formula = 
      `P1 Win` ~
      `Scored 1st` + 
      `P1 Shots` + `P2 Shots` +
      `P1 Hits` + `P2 Hits` +
      `P1 ToA` + `P2 ToA` +
      `P1 Passing` + `P2 Passing` + 
      `P1 Faceoffs` + `P2 Faceoffs` +
      `P1 Offensive Faceoffs` + `P2 Offensive Faceoffs` +
      `P1 PM` + `P2 PM` +
      `P1 Powerplays` + `P2 Powerplays` +
      `P1 Penalty Shots` + `P2 Penalty Shots` +
      `P1 Breakaways` + `P2 Breakaways`
    ,family = binomial
    ,data =   hockey_data
  )
coef(hockey_full_logit_02)
```
* Scored 1st is significant
* P1, P2 Passing, and P1 Breakaways all have p-values $<0.10$


## Backward BIC
```{r Stepwise 01, include=FALSE}
hockey_full_logit_02 %>% 
  step(
    direction = "backward"
    ,k = 
      hockey_data %>% 
      count() %>% 
      as.numeric() %>% 
      log()
    ,trace = 0
  ) %>% 
  summary()

hockey_step_logit_01 <- 
  hockey_data %>% 
  glm(
    `P1 Win` ~ `Scored 1st` + `P1 Powerplays` + `P2 Penalty Shots`
    ,family = binomial
    ,data = .
  )
goodness_of_fit(hockey_step_logit_01, type = "Chisq")
goodness_of_fit(hockey_step_logit_01, type = "Gsq")
```

$$
  \operatorname{Logit}\left( \text{P1 Wins} \right) =
  \alpha + \beta_{1}\text{Scored 1st} + \beta_{2}\text{P1 Powerplays} 
$$
$$
  + \beta_{3}\text{P2 Penalty Shots}
$$
$$
  =
  -0.8114 - 1.7359\text{Scored 1st} + 0.2133\text{P1 Powerplays}
$$
$$
  - 1.5155\text{P2 Penalty Shots}
$$
Where:
$$
  \text{Scored 1st} =
  \begin{cases}
    0,\text{ if P1} \\
    1,\text{ if P2}
  \end{cases}
$$

```{r Stepwise 01 CI}
hockey_step_logit_01 %>% 
  confint() %>% 
  exp()
```



# LASSO
```{r LASSO Variables, include=FALSE}
ind_var_01 <- 
  c(
    "Scored 1st Boolean (P2)"
    ,"P1 Shots"
    ,"P2 Shots"
    ,"P1 Hits"
    ,"P2 Hits"
    ,"P1 ToA"
    ,"P2 ToA"
    ,"P1 Passing"
    ,"P2 Passing"
    ,"P1 PM"
    ,"P2 PM"
    ,"P1 Powerplays"
    ,"P2 Powerplays"
    ,"P1 Faceoffs"
    ,"P2 Faceoffs"
    ,"P1 Breakaways"
    ,"P2 Breakaways"
    ,"P1 Penalty Shots"
    ,"P2 Penalty Shots"
    ,"P1 Offensive Faceoffs"
    ,"P2 Offensive Faceoffs"
  )
ind_var_01

dep_var <- "P1 Win"

hockey_lasso_ind_01 <- 
  hockey_data %>% 
  select(ind_var_01) %>% 
  as.matrix()

hockey_lasso_dep <-
  hockey_data %>% 
  select(dep_var) %>% 
  as.matrix()
```


## LASSO Plot
```{r LASSO Plot}
hockey_lasso_cvfit_01 <-
  cv.glmnet(
    x = hockey_lasso_ind_01
    ,y = hockey_lasso_dep
    ,family = "binomial"
    ,alpha = 1
  )

autoplot(hockey_lasso_cvfit_01)
```


## LASSO Coefficients 1se
```{r LASSO Coefficients 1se}
hockey_lasso_coef_01_1se <- coef(hockey_lasso_cvfit_01, s = "lambda.1se", exact = TRUE)
hockey_lasso_coef_01_1se
```


## LASSO Coefficients Min
```{r LASSO Coefficients min}
hockey_lasso_coef_01_min <- coef(hockey_lasso_cvfit_01, s = "lambda.min", exact = TRUE)
hockey_lasso_coef_01_min
```


## LASSO Average Coefficient
```{r LASSO Repeated Coefficient, cache=TRUE}
reps <- 100

hockey_lasso_res_01 <- matrix(nrow = length(ind_var_01) + 1, ncol = reps)
rownames(hockey_lasso_res_01) <- c("(Intercept)", ind_var_01)

hockey_lasso_lambda_01 <- tibble(Min = as.numeric(NA), SE = as.numeric(NA))

for(i in 1:reps){
  hockey_lasso_cvfit_rep <- 
    cv.glmnet(
      x = hockey_lasso_ind_01
      ,y = hockey_lasso_dep
      ,family = "binomial"
      ,alpha = 1
    )
  
  hockey_lasso_coef_rep_min <- 
    hockey_lasso_cvfit_rep %>% 
    coef(s = "lambda.min", exact = TRUE) %>% 
    as.matrix()
  hockey_lasso_res_01[, i] <- hockey_lasso_coef_rep_min
  hockey_lasso_lambda_01[i, "Min"] <- hockey_lasso_cvfit_rep$lambda.min
  hockey_lasso_lambda_01[i, "SE"] <- hockey_lasso_cvfit_rep$lambda.1se
}

apply(X = hockey_lasso_res_01, MARGIN = 1, FUN = mean)
```


## LASSO Average Lambda
```{r LASSO Repeated Lambda}
hockey_lasso_avg_lambda <- 
  cv.glmnet(
  x = hockey_lasso_ind_01
  ,y = hockey_lasso_dep
  ,family = "binomial"
  ,alpha = 1
) %>% 
  coef(s = mean(hockey_lasso_lambda_01$Min))
hockey_lasso_avg_lambda
```



# Results
## Model
- LASSO Average Lambda chosen
    - Scored First
    - P1 Shots
    - P1 \& P2 Passing
    - P1 Powerplays
    - P1 Breakaways
    - P2 Penalty Shots
    
$$
  \ln\left( \frac{\text{P1 Wins}}{\text{P2 Wins}} \right) =
  -2.2919 - 1.4114\text{Scored First} + 0.0046\text{P1 Shots}
$$
$$
  + 1.0339\text{P1 Passing} + 1.1588\text{P2 Passing} + 0.1212\text{P1 Powerplays} 
$$
$$
  + 0.0623\text{P1 Breakaways} - 0.5254\text{P2 Penalty Shots}
$$


## Interpretation
- By scoring first, a player increases their estimated odds in favour of winning by a multiplicative factor of `r round(exp(-hockey_lasso_avg_lambda["Scored 1st Boolean (P2)",]), 4)`

- Only P1's shots was found as a useful predictive measure.
    - Increases the estimated odds in favour by a multiplicative factor of `r round(exp(hockey_lasso_avg_lambda["P1 Shots",]), 4)` per each additional shot
    - So not a lot
    
- If both players make a lot of their passes, P1's increases their estimated odds in favour of winning by a multiplicative factor of:
    - P1: `r round(exp(hockey_lasso_avg_lambda["P1 Passing",]), 4)`
    - P2: `r round(exp(hockey_lasso_avg_lambda["P2 Passing",]), 4)`
    
- The more 1-on-0 opportunities given up in a game by a player, the less likely that player will win.
    - Breakaways: `r round(exp(hockey_lasso_avg_lambda["P1 Breakaways",]), 4)`
    - Penalty Shots: `r round(exp(hockey_lasso_avg_lambda["P2 Penalty Shots",]), 4)`


## Weaknesses
- Based on a video game
    - May not apply to real life

- Who is P1 in real life?
    - Other combinations of players may have different results

- Includes P1's and P2's Passing \%
    - Both have the same sign
    - Unexpected
    
- Possible overfitting
    - Uses $\lambda_{Min}$ instead of $\lambda_{1se}$
    
- Introduced bias to lower variance

- It may be beneficial to use the average coefficient LASSO, with a cutoff if the variable appears in $<x\%$ of results


## References
Gelman, A. (2014, June 02). Why we hate stepwise regression. Retrieved November 29, 2017, from http://andrewgelman.com/2014/06/02/hate-stepwise-regression/

Hastie, T., & Qian, J. (2014, June 26). Glmnet Vignette. Retrieved November 29, 2017, from https://web.stanford.edu/~hastie/glmnet/glmnet_alpha.html

NHL 18 - EA SPORTS - Official Site. (n.d.). Retrieved November 29, 2017, from https://www.easports.com/nhl

R. (n.d.). 2_buds_hockey_stats.xlsx. Retrieved November 29, 2017, from http://randude.com/ps3/

Testing for Autocorrelation. (1999). Retrieved November 29, 2017, from http://www.okstate.edu/sas/v8/sashtml/ets/chap8/sect5.htm

Wickham, H. (n.d.). Tidyverse. Retrieved November 29, 2017, from https://www.tidyverse.org/
