---
title: "Randomization Test"
author: "Scott Graham"
date: "December 06, 2017"
header-includes:
   - \usepackage{mathtools}
   - \DeclarePairedDelimiter\ceil{\lceil}{\rceil}
   - \DeclarePairedDelimiter\floor{\lfloor}{\rfloor}
output: 
  ioslides_presentation:
    smaller: true
    logo: ../../../images/logo/GrahamStat-emblem-black.png
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(tidyverse, warn.conflicts = FALSE, quietly = TRUE)
library(magrittr, warn.conflicts = FALSE, quietly = TRUE)
library(perm, warn.conflicts = FALSE, quietly = TRUE)
library(coin, warn.conflicts = FALSE, quietly = TRUE)
library(knitr, warn.conflicts = FALSE, quietly = TRUE)
library(kableExtra, warn.conflicts = FALSE, quietly = TRUE)

theme_minimal2 <- theme_minimal() %>%  theme_set()
theme_minimal2 <-
  theme_update(
    panel.border = element_rect(
      linetype = "solid"
      ,colour = "grey92"
      ,fill = NA
    )
    ,strip.background = element_rect(
      linetype = "solid"
      ,colour = "grey92"
      ,fill = NA
    )
  )

four_facet_hist <- function(data){
  data %>%
    gather(key = X_i, value = Observation, -j) %>%
    ggplot(
      aes(
        x = Observation,
        colour = X_i
      )
    ) +
    geom_density(size = 1) +
    facet_wrap(
      ~ X_i
      ,ncol = 1
    ) +
    scale_colour_brewer(
      type = "qual"
      ,name = "X_i"
      ,palette = "Set2"
    )
}

permPT <- function(x, y, alpha){
  # D_i is their difference
  D_i <- y - x

  # Remove any 0s
  D_i <- D_i[D_i != 0]
  n <- length(D_i)
  
  # Test Statistic
  T_2 <- sum(D_i[D_i > 0])
  
  # Quantiles
  lower_quantile <- ceiling(2^n * alpha / 2)
  
  # Orders the D_i from smallest to largest
  D_i <- D_i[order(D_i)]
  
  # Creates a matrix of size n, 2^n
  all_possible_combs <- t(expand.grid(lapply(numeric(n), function(x) c(-1, 1))))
  all_possible_combs <- all_possible_combs * D_i
  
  # Anything <= 0, we set to 0, so we don't sum them when finding T_2
  for (i in 1:n){
    for(j in 1:(2^n)){
      all_possible_combs[i, j] <- ifelse(all_possible_combs[i, j] > 0, all_possible_combs[i, j], 0)
    }
  }
  
  # Finds the column sums, orders them and prints out the required ones
  all_possible_col_sums <- colSums(all_possible_combs)
  all_possible_col_sums <- all_possible_col_sums[order(all_possible_col_sums)]
  
  # Rejection Region
  w_lower <- all_possible_col_sums[lower_quantile]
  w_upper <- sum(abs(D_i)) - w_lower
  RR <- c(w_lower, w_upper)
  
  # P-Value
  num_leq_T_2 <- length(all_possible_col_sums[all_possible_col_sums <= T_2])
  num_geq_T_2 <- length(all_possible_col_sums[all_possible_col_sums >= T_2])
  p.value = 2 * (min(num_leq_T_2, num_geq_T_2) / 2^(n))
  
  # Results List
  results <- 
    list(
      T_2 = T_2
      ,RR = RR
      ,p.value = p.value
    )
  results
}

permPT_time <- function(iterations){
  function_time_df <- 
    tibble(
      n = as.numeric(NA)
      ,Time = as.numeric(NA)
    )
  for(i in 2:iterations){
    x <- rnorm(n = i, mean = 0, sd = 1)
    y <- rnorm(n = i, mean = 1, sd = 1)
    function_time_df[i-1, ] <- 
      cbind(
        i
        ,as.numeric(system.time(expr = permPT(x, y, alpha = 0.05))["elapsed"])
      )
  }
  rownames(function_time_df) <- NULL
  function_time_df
}
```
\[
  \newcommand{\Prob}{\operatorname{P}}
  \newcommand{\E}{\operatorname{E}}
  \newcommand{\Var}{\operatorname{Var}}
  \newcommand{\Cov}{\operatorname{Cov}}
  \newcommand{\se}{\operatorname{se}}
  \newcommand{\re}{\operatorname{re}}
  \newcommand{\ybar}{{\overline{Y}}}
  \newcommand{\phat}{{\hat{p}}}
  \newcommand{\that}{{\hat{T}}}
  \newcommand{\med}{{\tilde{Y}}}
\]

# Theory
## Motivation:
- Hard to determine underlying distribution of data and statistics

- Lose relative efficiency with Ranks

- Can we use the observations?

- Created by Fisher in 1935

- Also known as Permutation Tests

$$
  H_{0}: \theta(X) = \theta(Y)
$$
$$
  H_{1}: \theta(X) \neq \theta(Y)
$$

- $\theta$ can be any number of statistics, with either known, or unknown underlying distributions.



# Independent Samples
## Data:
- Consider two independent simple random samples:
$$
  X_{i}, i = 1, 2, \dots, n
$$
$$
  Y_{j}, j = 1, 2, \dots, m
$$

- Where:
$$
  X_{i} \stackrel{i.i.d.}{\sim} f(x) \forall i = 1, 2, \dots, n
$$
$$
  Y_{j} \stackrel{i.i.d.}{\sim} g(y) \forall j = 1, 2, \dots, m
$$

- Both $X$ and $Y$ are at least an interval scale

- We then pool the data into one large random sample $Z$
$$
  Z = 
  \{X_{i}, Y_{j}\} \forall i,j \implies
  Z = 
  \{X_{1}, X_{2}, \dots, X_{n}, Y_{1}, Y_{2}, \dots, Y_{m}\}
$$


## Test Statistic:
Let:
$$
  T_{1} =
  \sum_{i = 1}^{n} X_{i}
$$


## Distribution:
- We consider all ways to choose $n$ observations from $Z$, of which there are $n+m$ entries

- We have exactly ${{n+m}\choose{n}}$ possible combinations

- Then we find the arrangement of $X$ and $Y$ that yield the smallest $T_{1}$

- Order them from smallest to largest, we use:
$$
  p^{th}-quantile =
  \left\lceil {{n+m}\choose{n}}p \right\rceil
$$

- This number is the index of the value of $T_{1}$ we wish to assign to $\omega_{p}$


## Distribution:
- Let $T_{1}^{(i)}$ be the $i^{\text{th}}$ largest possible value for $T_{1}$, by rearranging which values in $Z$ are considered as $X$s or $Y$s

- Then we have the pth-quantile of $T_{1}$:
$$
  \omega_{p} =
  T_{1}^{\left( \left\lceil {{n+m}\choose{n}}p \right\rceil \right)}
$$

- Under the null hypothesis, approximately $p$ percent of $T_{1}$s should be less than $T_{1}^{(\omega_{p})}$


## Testing:
- Rejection Region:
    - $\omega_{\frac{\alpha}{2}}$
    - $\omega_{1-\frac{\alpha}{2}}$
    - See if $T_{1}$ falls outside of these two numbers.

- P-Value:
    - Find the proportion of possible values for $T_{1}$, less than what we observed
    - Multiply it by 2 if we are considering the two-tailed version.


## Proof of Concept (Manual):
$$
  X = \{5, 1, 2\},
  Y = \{6, 5, 5, 4\} \implies
  Z = \{1, 2, 4, 5, 5, 5, 6\}
$$
$$
  H_{0}: \E(X) = \E(Y)
$$
$$
  H_{1}: \E(X) \neq \E(Y)
$$
$$
  \frac{\alpha}{2}-percentile =
  \left\lceil {{3+4}\choose{3}}\frac{0.05}{2} \right\rceil =
  \left\lceil 0.875 \right\rceil =
  1
$$
$$
  1-\frac{\alpha}{2}-percentile =
  \left\lceil {{3+4}\choose{3}}\left( 1-\frac{0.05}{2} \right) \right\rceil =
  \left\lceil 34.25 \right\rceil =
  35
$$


## Possible Combinations:
|    (i)   |     1    |     2    |     4    |     5    |     5    |     5    |     6    |     T    |
|:--------:|:--------:|:--------:|:--------:|:--------:|:--------:|:--------:|:--------:|:--------:|
|    (1)   |     X    |     X    |     X    |     Y    |     Y    |     Y    |     Y    |     7    |
|    (2)   |     X    |     X    |     Y    |     X    |     Y    |     Y    |     Y    |     8    |
|    (3)   |     X    |     X    |     Y    |     Y    |     X    |     Y    |     X    |     8    |
| $\vdots$ | $\vdots$ | $\vdots$ | $\vdots$ | $\vdots$ | $\vdots$ | $\vdots$ | $\vdots$ | $\vdots$ |
|   (34)   |     Y    |     Y    |     Y    |     X    |     Y    |     X    |     X    |    16    |
|   (35)   |     Y    |     Y    |     Y    |     Y    |     X    |     X    |     X    |    16    |

$$
  T_{1} =
  \sum_{i=1}^{3}X_{i} =
  5 + 1 + 2 =
  13
$$

## Results:
- Therefore we fail to reject the null hypothesis based on the sample.
$$
  p-value =
  2\frac{4}{35} =
  \frac{8}{35} =
  0.2286
$$

- We use 4 in the numerator, as there are 4 possible values for $T_{1}\leq13$.

## Proof of Concept (R):
- This is equivalent to doing the following:
- `library(perm)`
- the function `permTS`

```{r PoCIS, echo=TRUE}
set.seed(5609)
x_PoC <- sample(1:5, size = 3, replace = TRUE)
y_PoC <- sample(3:8, size = 4, replace = TRUE)
perm::permTS(x = x_PoC, y = y_PoC, alternative = "two.sided", exact = TRUE)
```



# Matched Pairs
## Data:
- Another useful configuration is when $X$ and $Y$ are paired:
$$
  \left( X_{1}, Y_{1} \right), \left( X_{2}, Y_{2} \right), \dots, \left( X_{n'}, Y_{n'} \right)
$$


## Test Statistic:
- Only look at the $n$ pairs who have non zero differences. Let:
$$
  D_{i} = 
  Y_{i} - X_{i}, i = 1, 2, \dots, n
$$
$$
  T_{2} =
  \sum_{i=1}^{n}D_{i}, \forall D_{i} > 0
$$

- Considering $|D_{i}|$, there are $2^{n}$ ways of assigning a $+$ or $-$ symbol to each of the $n$ $D_{i}$s

- Follow a similar procedure to the independent sample method


## Distribution:
- We can define $\omega_{\frac{\alpha}{2}}$ as:
$$
  \frac{\alpha}{2}-quantile = 
  2^{n}\frac{\alpha}{2}
$$

- Let:
$$
  \omega_{\frac{\alpha}{2}} =
  T_{2}^{\left( 2^{n}\frac{\alpha}{2} \right)}
$$
$$
  \omega_{1 - \frac{\alpha}{2}} =
  \sum_{i=1}^{n}|D_{i}| - \omega_{\frac{\alpha}{2}}
$$

- Which forms the basis for the rejection region.

- The p-value for this test is:
$$
  2\frac{\min\{ \text{\# of }T_{2} \leq \text{the observed}, \text{\# of }T_{2} \geq \text{the observed} \}}{2^{n}}
$$


## Proof of Concept (Code):
```{r permPT Code 01, echo=TRUE, eval=FALSE}
permPT <- function(x, y, alpha){
  # D_i is their difference
  D_i <- y - x

  # Remove any 0s
  D_i <- D_i[D_i != 0]
  n <- length(D_i)
  
  # Test Statistic
  T_2 <- sum(D_i[D_i > 0])
  
  # Quantiles
  lower_quantile <- ceiling(2^n * alpha / 2)
  
  # Orders the D_i from smallest to largest
  D_i <- D_i[order(D_i)]
  
  # Creates a matrix of size n, 2^n
  all_possible_combs <- t(expand.grid(lapply(numeric(n), function(x) c(-1, 1))))
  all_possible_combs <- all_possible_combs * D_i
```


## Proof of Concept (Code):
```{r permPT Code 02, echo=TRUE, eval=FALSE}
  # Anything <= 0, we set to 0, so we don't sum them when finding T_2
  for (i in 1:n){
    for(j in 1:(2^n)){
      all_possible_combs[i, j] <- 
        ifelse(all_possible_combs[i, j] > 0, all_possible_combs[i, j], 0)
    }
  }
  
  # Finds the column sums, orders them and prints out the required ones
  all_possible_col_sums <- colSums(all_possible_combs)
  all_possible_col_sums <- all_possible_col_sums[order(all_possible_col_sums)]
  
  # Rejection Region
  w_lower <- all_possible_col_sums[lower_quantile]
  w_upper <- sum(abs(D_i)) - w_lower
  RR <- c(w_lower, w_upper)
  
  # P-Value
  num_leq_T_2 <- length(all_possible_col_sums[all_possible_col_sums <= T_2])
  num_geq_T_2 <- length(all_possible_col_sums[all_possible_col_sums >= T_2])
  p.value = 2 * (min(num_leq_T_2, num_geq_T_2) / 2^(n))
```


## Proof of Concept (Code):
```{r permPT Code 03, echo=TRUE, eval=FALSE}
  # Results List
  results <- 
    list(
      T_2 = T_2
      ,RR = RR
      ,p.value = p.value
    )
  results
}
```


## Proof of Concept (R):
### Data:
```{r PoCTS 01, echo=TRUE}
set.seed(5609)
alpha <- 0.05
n_prime <- 20
# Random sample with n' = 10
z_PoC <- sample(-5:5, size = n_prime, replace = TRUE)
z_PoC

# First n'/2 numbers are assigned to x, second n'/2 to y
x_PoC <- z_PoC[1:(n_prime/2)]
y_PoC <- z_PoC[(n_prime/2 + 1):n_prime]

results <- permPT(x_PoC, y_PoC, alpha)
```


## Proof of Concept (R):
### Results:
```{r PoCTS 02}
cat(
  "\n",
  "T_2:", results$T_2, "\n",
  "Lower Bound:", results$RR[1], "\n",
  "Upper Bound:", results$RR[2], "\n",
  "P-Value:", results$p.value
)
```

- Since $T_{2}=8\in[2, 26]$, and equivalently, the p-value$=0.40625>\alpha=0.05$, we fail to reject the null hypothesis of equal means based on the sample.


## Time to Run:
```{r permPT Timing, cache=TRUE}
permPT_time(15) %>% 
  ggplot(
    aes(
      x = n
      ,y = Time
    )
  ) +
  geom_smooth(
    method = "lm"
    ,formula = y ~ exp(x)
    ,se = FALSE
  ) +
  geom_point() +
  labs(y = "Time (s)")
```



# Conclusion:
## Results:
- Effectively allows you to test a hypothesis involving nearly any statistic, without the need to derive an underlying distribution

- By freeing oneself of this need, the assumptions for running such a test are relaxed
    - This can be especially helpful for cases with $n<20$, as when examining the mean, the Central Limit Theorem may not have kicked in yet

- A drawback of this method is it is typically not the optimal solution in most cases

- Unless carefully written, Randomization tests can be quite computationally intensive 
    - `permPT`


## References
Conover, W. J. (1999). Practical nonparametric statistics (3rd ed.). John Wiley & Sons.

Fay, M. (2010, July 29). Perm: Exact or Asymptotic permutation tests. Retrieved November 29, 2017, from https://cran.r-project.org/package=perm